---
title: "FoodPricing2019_NeighborhoodCharacteristics"
author: "Aldo Crossa"
date: "2023-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Survey of food prices in NYC, 2019: Neighborhood Characteristics

This code was writtedn to generate neighborhood-level estimates of various socioeconomic characteristics. This script pulls data from the 2015-2019 American community Survey, using the Census API.

We begin by loading the necessary packages

```{r Load packages}
###load packages###
library(censusapi)
library(data.table)
library(dplyr)
library(stringr)
library(sqldf)
library(xlsx)

setwd("<<insert path here>>")
###API key (sign up for a key here: https://api.census.gov/data/key_signup.html)
mycensuskey <-"<<insert census key here>>"


```

## ACS variables to pull from the census API 

Here we define what variables to pull from the Census API in order to calculate the neighborhood level indicators. Define the labels for each variable and group the variables.

```{r Variables to pull, echo=FALSE}

#keep labels for all variables that will be used in the neighborhood-level indicators

varlabels <- c("B25001_001E" , "Total (For housing Units)", 
"B25003_001E" , "Total (For housing units by tenure)", 
"B25003_002E" , "Total: Owner occupied", 
"B25106_024E" , "Total: Renter-occupied housing units:", 
"B25106_028E" , "Total: Renter-occupied housing units: Less than $20,000: 30 percent or more", 
"B25106_032E" , "Total: Renter-occupied housing units: $20,000 to $34,999: 30 percent or more", 
"B25106_036E" , "Total: Renter-occupied housing units: $35,000 to $49,999: 30 percent or more", 
"B25106_040E" , "Total: Renter-occupied housing units: $50,000 to $74,999: 30 percent or more", 
"B25106_044E" , "Total: Renter-occupied housing units: $75,000 or more: 30 percent or more", 
"B25106_045E" , "Total: Renter-occupied housing units: Zero or negative income", 
"B25077_001E" , "Median value (dollars)", 
"B25076_001E" , "Lower value quartile (dollars)", 
"B25078_001E" , "Upper value quartile (dollars)", 
"B22001_001E" , "Total: (For SNAP)", 
"B22001_002E" , "Total: Household received Food Stamps/SNAP in the past 12 months:", 
"C17002_001E" , "Total (for poverty)", 
"C17002_002E" , "Total: Ratio income to poverty level, Under .50", 
"C17002_003E" , "Total: Ratio income to poverty level, .50 to .99", 
"B23025_001E" , "Total (for employment, population 16+)", 
"B23025_003E" , "Total: In labor force: Civilian labor force:", 
"B23025_004E" , "Total: In labor force: Civilian labor force: Employed", 
"B06009_001E" , "Total (For educational attainment)", 
"B06009_002E" , "Total: Less than high school graduate", 
"B16005_001E" , "Total (For language)", 
"B16005_007E" , "Total>Native>Speak Spanish>Speak English \"not well\"", 
"B16005_008E" , "Total>Native>Speak Spanish>Speak English \"not at all\"", 
"B16005_012E" , "Total>Native>Speak other Indo-European languages>Speak English \"not well\"", 
"B16005_013E" , "Total>Native>Speak other Indo-European languages>Speak English \"not at all\"", 
"B16005_017E" , "Total>Native>Speak Asian and Pacific Island languages>Speak English \"not well\"", 
"B16005_018E" , "Total>Native>Speak Asian and Pacific Island languages>Speak English \"not at all\"", 
"B16005_022E" , "Total>Native>Speak other languages>Speak English \"not well\"", 
"B16005_023E" , "Total>Native>Speak other languages>Speak English \"not at all\"", 
"B16005_029E" , "Total>Foreign born>Speak Spanish>Speak English \"not well\"", 
"B16005_030E" , "Total>Foreign born>Speak Spanish>Speak English \"not at all\"", 
"B16005_034E" , "Total>Foreign born>Speak other Indo-European languages>Speak English \"not well\"", 
"B16005_035E" , "Total>Foreign born>Speak other Indo-European languages>Speak English \"not at all\"", 
"B16005_039E" , "Total>Foreign born>Speak Asian and Pacific Island languages>Speak English \"not well\"", 
"B16005_040E" , "Total>Foreign born>Speak Asian and Pacific Island languages>Speak English \"not at all\"", 
"B16005_044E" , "Total>Foreign born>Speak other languages>Speak English \"not well\"", 
"B16005_045E" , "Total>Foreign born>Speak other languages>Speak English \"not at all\"", 
"B02001_001E" , "Total (for race)", 
"B02009_001E" , "Total:Black or African American", 
"B03001_001E" , "Total: (for Hispanic)", 
"B03001_003E" , "Total: Hispanic or Latino:", 
"B02011_001E" , "Total: Asian ", 
"B02013_001E" , "Total: Some other race", 
"B27010_002E" , "Total: Under 19 years: ", 
"B27010_017E" , "Total: Under 19 years: No health insurance coverage", 
"B27010_018E" , "Total: 19 to 34 years: ", 
"B27010_033E" , "Total: 19 to 34 years: No health insurance coverage", 
"B27010_034E" , "Total: 35 to 64 years: ", 
"B27010_050E" , "Total: 35 to 64 years: No health insurance coverage", 
"B27010_051E" , "Total: 65 years and over: ", 
"B27010_066E" , "Total: 65 years and over: No health insurance coverage", 
"B19083_001E" , "Gini Index of income inequality")

varattr <- as.data.frame(matrix(varlabels, byrow = 2, ncol = 2))
names(varattr) <- c("Variable.Name", "Variable.label")

AllVars <- varattr$Variable.Name

#### Group variables to use later in the calcultion of the indicators

# % Owner occupied
homevars <- c("B25001_001E","B25003_002E")

 
# % Rent burdened

rentvars <- c("B25106_028E","B25106_032E","B25106_036E","B25106_040E","B25106_044E","B25106_024E")


# % Household value

HHvars = "B25077_001E"


# % Cash Assistanceship or SNAP

snapvars = c("B22001_002E", "B22001_001E")

 
#   % Poverty 

povertyvars <- c("C17002_002E","C17002_003E","C17002_001E")

# % Employment Status (out of the entire population, not just in the labor force)

employvars <- c("B23025_004E", "B23025_001E")

 
# % Educational attainment

educvars <- c("B06009_002E","B06009_001E")


# % linguistic isolation?? Use "Ability to speak english"

langvars <- c("B16005_007E", "B16005_008E", "B16005_012E", "B16005_013E", "B16005_017E",
               "B16005_018E", "B16005_022E", "B16005_023E", "B16005_029E", "B16005_030E",
               "B16005_034E", "B16005_035E", "B16005_039E", "B16005_040E", "B16005_044E",
               "B16005_045E", "B16005_001E")

# %%%% % Black/Black American 

pblackvars <- c("B02009_001E","B02001_001E")
 
# %%%% % Hispanic/Latino

platinovars <- c("B03001_003E", "B03001_001E")


paapivar <- c("B02011_001E", "B02001_001E")


# %%%% Other

pothervars <- c("B02013_001E", "B02001_001E")

# % No health insurance

insrvars <- c("B27010_002E","B27010_018E","B27010_034E","B27010_051E",
              "B27010_017E","B27010_033E","B27010_050E","B27010_066E")


# % Gini Index

GINIvars <- "B19083_001E"

```

#NYC ZIP codes 

These are the ZIP codes for NYC

```{r NYC ZIP codes, echo=FALSE}

ZCTA_2010 <- c("10001", "10002", "10003", "10004", "10005", "10006", "10007", "10009", "10010", "10011", "10012", 
               "10013", "10014", "10016", "10017", "10018", "10019", "10020", "10021", "10022", "10023", "10024", 
               "10025", "10026", "10027", "10028", "10029", "10030", "10031", "10032", "10033", "10034", "10035",
               "10036", "10037", "10038", "10039", "10040", "10044", "10065", "10069", "10075", "10103", "10110", 
               "10111", "10112", "10115", "10119", "10128", "10152", "10153", "10154", "10162", "10165", "10167", 
               "10168", "10169", "10170", "10171", "10172", "10173", "10174", "10177", "10199", "10271", "10278", 
               "10279", "10280", "10282", "10301", "10302", "10303", "10304", "10305", "10306", "10307", "10308", 
               "10309", "10310", "10311", "10312", "10314", "10451", "10452", "10453", "10454", "10455", "10456", 
               "10457", "10458", "10459", "10460", "10461", "10462", "10463", "10464", "10465", "10466", "10467", 
               "10468", "10469", "10470", "10471", "10472", "10473", "10474", "10475", "11001", "11003", "11004", 
               "11005", "11040", "11101", "11102", "11103", "11104", "11105", "11106", "11109", "11201", "11203", 
               "11204", "11205", "11206", "11207", "11208", "11209", "11210", "11211", "11212", "11213", "11214", 
               "11215", "11216", "11217", "11218", "11219", "11220", "11221", "11222", "11223", "11224", "11225", 
               "11226", "11228", "11229", "11230", "11231", "11232", "11233", "11234", "11235", "11236", "11237", 
               "11238", "11239", "11351", "11354", "11355", "11356", "11357", "11358", "11359", "11360", "11361", 
               "11362", "11363", "11364", "11365", "11366", "11367", "11368", "11369", "11370", "11371", "11372",
               "11373", "11374", "11375", "11377", "11378", "11379", "11385", "11411", "11412", "11413", "11414", 
               "11415", "11416", "11417", "11418", "11419", "11420", "11421", "11422", "11423", "11424", "11425", 
               "11426", "11427", "11428", "11429", "11430", "11432", "11433", "11434", "11435", "11436", "11451", 
               "11691", "11692", "11693", "11694", "11697", "99999")

zcta.string <- paste(ZCTA_2010, collapse = ",")

```

#Next, pull the data using the Census API

Data will be pulled from the ACS 2015-2019 5-year estimates using the "getCensus" function within the censusapi package. 


```{r Pull data, echo= FALSE}

# Pull the data from the API
acsdata5y_zcta <- getCensus(name = "acs/acs5",
                                     vintage = 2019,
                                     key = mycensuskey,
                                     vars = AllVars,
                                     region = paste0("zip code tabulation area:",zcta.string),
                                     regionin = "state:36")


# Add labels to the variables 
for (i in 1:dim(varattr)[1]){
    setattr( acsdata5y_zcta[, varattr$Variable.Name[i]], 
             "label", 
             varattr$Variable.label[i])
}


```


# Calculating Neighborhood level indicators

```{r Get Indicators, echo = FALSE}

#### Create Indicators ----
acsdata5y_zcta <-  
  subset(acsdata5y_zcta, subset = (B02001_001E > 0 & !(zip_code_tabulation_area %in% c(10162, 10279, 11430)))) %>%
            mutate( percent_owner = B25003_002E/B25001_001E,
                    percent_rentburdened = (B25106_028E+B25106_032E+B25106_036E+B25106_040E+B25106_044E)/B25106_024E,
                    HH_value = case_when(B25077_001E >= 0 ~ B25077_001E),
                    percent_snap = B22001_002E / B22001_001E,
                    percent_poverty = (C17002_002E+C17002_003E)/C17002_001E,
                    percent_employed = B23025_004E / B23025_001E,
                    percent_lths = B06009_002E/B06009_001E,
                    percent_noenglish = 
                      (B16005_007E + B16005_008E + B16005_012E + B16005_013E + B16005_017E +
                         B16005_018E + B16005_022E + B16005_023E + B16005_029E + B16005_030E +
                         B16005_034E + B16005_035E + B16005_039E + B16005_040E + B16005_044E +
                         B16005_045E) / B16005_001E,
                    percent_black = B02009_001E/B02001_001E,
                    percent_latino = B03001_003E/B03001_001E,
                    percent_aapi = B02011_001E/B02001_001E,
                    percent_other = B02013_001E/B02001_001E,
                    percent_noinsurance = (B27010_017E+B27010_033E+B27010_050E+B27010_066E)/
                                          (B27010_002E+B27010_018E+B27010_034E+B27010_051E),
                    GINI_index = case_when(B19083_001E > 0 ~ B19083_001E)
            )

#### Set labels for new variables ----
setattr( acsdata5y_zcta$percent_owner, "label", "% Owner occupied")
setattr( acsdata5y_zcta$percent_rentburdened, "label", "% Rent burdened")
setattr( acsdata5y_zcta$HH_value, "label", "Median home value")
setattr( acsdata5y_zcta$percent_snap, "label", "% SNAP recipients")
setattr( acsdata5y_zcta$percent_poverty, "label", "% Living in poverty")
setattr( acsdata5y_zcta$percent_employed, "label", "% Employed")
setattr( acsdata5y_zcta$percent_lths, "label", "% Less than high school")
setattr( acsdata5y_zcta$percent_noenglish, "label", "% Not english speaker")
setattr( acsdata5y_zcta$percent_black, "label", "% Black or African American")
setattr( acsdata5y_zcta$percent_latino, "label", "% Latino")
setattr( acsdata5y_zcta$percent_aapi, "label", "% AAPI")
setattr( acsdata5y_zcta$percent_other, "label", "% Other Race")
setattr( acsdata5y_zcta$percent_noinsurance, "label", "% No insurance")
setattr( acsdata5y_zcta$GINI_index, "label", "GINI index of income inequality")


```


#Categorizing indicators

```{r Categorizing indicators, echo = FALSE}

#### Generate quartiles for each indicator ----
acsdata5y_zcta <-  
  acsdata5y_zcta %>%
  mutate( qrt_percent_owner = cut(100*percent_owner, 
                                  quantile(100*percent_owner, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_rentburdened = cut(100*percent_rentburdened,
                                         quantile(100*percent_rentburdened, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_HH_value = cut(HH_value,
                             quantile(HH_value, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_snap = cut(100*percent_snap,
                                 quantile(100*percent_snap, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_poverty = cut(100*percent_poverty,
                                    quantile(100*percent_poverty, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_employed = cut(100*percent_employed,
                                     quantile(100*percent_employed, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_lths = cut(100*percent_lths,
                                 quantile(100*percent_lths, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_noenglish = cut(100*percent_noenglish,
                                      quantile(100*percent_noenglish, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_black = cut(100*percent_black,
                                  quantile(100*percent_black, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_latino = cut(100*percent_latino,
                                   quantile(100*percent_latino, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_aapi = cut(100*percent_aapi,
                                 quantile(100*percent_aapi, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_other = cut(100*percent_other,
                                  quantile(100*percent_other, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_percent_noinsurance = cut(100*percent_noinsurance,
                                        quantile(100*percent_noinsurance, probs =seq(0,1,0.25), na.rm = TRUE, names = T)),
          qrt_GINI_index = cut(GINI_index,
                                quantile(GINI_index, probs =seq(0,1,0.25), na.rm = TRUE, names = T))
  )

#### Create labels for categorized indicators
setattr( acsdata5y_zcta$qrt_percent_owner, "label", "(Quartiles) % Owner occupied")
setattr( acsdata5y_zcta$qrt_percent_rentburdened, "label", "(Quartiles) % Rent burdened")
setattr( acsdata5y_zcta$qrt_HH_value, "label", "(Quartiles) Median home value")
setattr( acsdata5y_zcta$qrt_percent_snap, "label", "(Quartiles) % SNAP recipients")
setattr( acsdata5y_zcta$qrt_percent_poverty, "label", "(Quartiles) % Living in poverty")
setattr( acsdata5y_zcta$qrt_percent_employed, "label", "(Quartiles) % Employed")
setattr( acsdata5y_zcta$qrt_percent_lths, "label", "(Quartiles) % Less than high school")
setattr( acsdata5y_zcta$qrt_percent_noenglish, "label", "(Quartiles) % Not english speaker")
setattr( acsdata5y_zcta$qrt_percent_black, "label", "(Quartiles) % Black or African American")
setattr( acsdata5y_zcta$qrt_percent_latino, "label", "(Quartiles) % Latino")
setattr( acsdata5y_zcta$qrt_percent_aapi, "label", "(Quartiles) % AAPI")
setattr( acsdata5y_zcta$qrt_percent_other, "label", "(Quartiles) % Other Race")
setattr( acsdata5y_zcta$qrt_percent_noinsurance, "label", "(Quartiles) % No insurance")
setattr( acsdata5y_zcta$qrt_GINI_index, "label", "(Quartiles) GINI Index of income inequality")


```


